% acronyms
\newacronym{api}{API}{Application Programming Interface}
\newacronym{hpc}{HPC}{high-performance computing}
\newacronym{tdd}{TDD}{test-driven development}
\newacronym{xml}{XML}{Extensible Markup Language}

% regular terms
\newglossaryentry{goroutine}{
    name={goroutine},
    description={A lightweigth concurrently executing function which gets multiplexed into OS threads by the Go runtime~\cite{effective_go}},
    plural={goroutines}
}
\newglossaryentry{intrinsic}{
    name={intrinsic},
    description={An intrinsic function is a function in a programming language which is handled specifically by the compiler. This is often used to optimize common operations for the target processor.}
}
\newglossaryentry{iterator_invalidation}{
    name={iterator invalidation},
    description={A common problem in languages without automatic memory management which can occur when an iterator is used on a mutable container. For example when iterating over a dynamically growing vector which reallocates itself, the pending iterator pointer can become dangling. Thus making it effectively unusable or invalid.}
}
\newglossaryentry{llvm}{
    name={LLVM},
    description={The LLVM Compiler Infrastructure Project (formerly short for Low Level Virtual Machine) is an umbrella project for various compiler and other low-level tools. LLVM Core is the primary subproject and a set of libraries for code generation and optimization for various platforms.~\cite{llvm_main}}
}
\newglossaryentry{loop_unrolling}{
    name={loop unrolling},
    description={Loop unrolling or loop unwinding is a common optimization used by compilers trading binary-size for speed. A loop gets transformed into the seperate instructions to avoid the overhead of the loop control instructions.}
}
\newglossaryentry{protobuf}{
    name={Protocol Buffers},
    description={``Protocol buffers are Google's language-neutral, platform-neutral, extensible mechanism for serializing structured data''~\cite{protobuf}}
}

% dual entries
\newdualentry
{beam}
{BEAM}
{Bogdan/Bj\"orn's Erlang Abstract Machine}
{The virtual machine which runs Erlang. It loads bytecode which is converted directly to threaded native code and executed.}
\newdualentry
{llvm_ir}
{LLVM IR}
{LLVM Intermediate Representation}
{A low level programming language similar to assembly. It is the code representation LLVM uses in its Core libraries. LLVM IR is platform-agnostic with the ``capability of representing `all` high-level languages cleanly''~\cite{llvm_ir_ref}.}
\newdualentry
{lts}
{LTS}
{Long Term Support}
{Mostly applied to Linux distributions, a Long Time Support version is a specific release with an extended support cycle. This means no new feaures get added but security patches are backported. LTS versions therefore offer better stability and security for mission-critical systems.}
\newdualentry
{mpi}
{MPI}
{Message Passing Interface}
{The Message Passing Interface standard is a library specification developed by a committee of vendors, implementors and users. It is the current cominant model used in high-performance computing and implementations for many platforms (both commercial and free) are available including bindings for various programming languages~\cite{mpi_main, mpi_infiniband}.}
\newdualentry
{osi}
{OSI}
{Open Source Initiative}
{The Open Source Initiative is a non-profit organization based in California which was ``formed to educate about and advocate for the benefits of open source and to build bridges among different constituencies in the open source community''~\cite{osi_about}.}
\newdualentry
{osm}
{OSM}
{OpenStreetMap}
{OpenStreetMap is a collaborative project which aims collect and maintain geographical data about roads, trails, railways stations and more. As the name suggests the data is provided openly under the Open Data Commons Open Database License~\fnote{\url{http://opendatacommons.org/licenses/odbl/}}.}
\newdualentry
{sloc}
{SLOC}
{source lines of code}
{A software metric counting the number of lines in a program's source code. Often used to roughly estimate developer productivity and maintainability}
\newdualentry
{ssh}
{SSH}
{Secure Shell}
{``Secure Shell is a cryptographic (encrypted) network protocol for initiating text-based shell sessions on remote machines in a secure way''~\cite{wiki_ssh}.}
\newdualentry
{sssp}
{SSSP}
{Single Source Shortest Path}
{A common graph problem searching for shortest paths between nodes of a graph. As the name suggests this type of problem states the use a single node as starting point and aims to determine shortest paths to all remaining nodes of the graph. Dijkstra's algorithm is commonly used for graphs with nonnegative edge weights while the Bellman-Ford-Algorithm can even handle that case.}
